---
title: "***DSCI605_M5_LAB4.BUBBLE AND BOX PLOT***"
author: "***CHARLES WIREDU***"
date: "2/8/2022"
output: pdf_document
fontsize: 11pt
mainfont: Calibri Light
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("~/Desktop/Data science Assignments")

### install packages required for the plots
##install.packages("ggplot2")
##install.packages("dplyr")
##install.packages("viridis")

library(ggplot2)
library(dplyr)
library(viridis)
```


```{r}
######load dataset for bubble graph
###################USArrests##################################
data("USArrests")
df <- USArrests
names(USArrests)

USArrests_data <- 
  tibble(State = state.name,
         Region = state.region) %>%
  bind_cols(USArrests)

names(USArrests_data)

```
---
The data set used is on the arrests in the United States in relation to 50 states. 
The data set has four variables namely, murder, assault, rape and urban pop. 
These variables were chosen because of their numeric description that will allow 
to demonstrate its  relationships.
---
```{r}
##############basic bubble plot######################
###PLOT1###
USArrests_data %>% 
  ggplot(aes(x=Assault,y=Murder, size=Rape)) +
  geom_point(alpha=0.6)+
  scale_size_continuous(range = c(1, 8))+
  labs(x="Assault", y="Murder", size="Rape")

##adding more features1
USArrests_data %>%
  arrange(desc(UrbanPop)) %>%
  ggplot(aes(x=Murder, y=Assault, size=Rape ,fill=Region)) +
  geom_point(alpha=0.5, shape=21, color="black") +
  scale_size(range = c(.1, 7), name="Urban Population ") +
  scale_fill_viridis(discrete=TRUE, guide=FALSE, option="A") +
  theme(legend.position="bottom")+
  ylab("Murder") +
  xlab("Assault") +
  theme(legend.position = "none")

##Graphs builds  bubble plot for murder and rape, murder and assault,couple with
##the display of rape, urban pop variable. From the graph there is a positive 
##relationship between x (Assault)and y (Murder) variable since all the 
## bubbles skew towards the right

###features 2 with title added
USArrests_data %>% 
  ggplot(aes(x=Rape,y=Murder, size=UrbanPop, color=Assault)) + 
  geom_point(alpha=0.8)+
  scale_size_continuous(range = c(1, 8))+
  labs(title="Bubble plot on Rape/Murder in Usa", x="Rape", y="Murder", size="UrbanPop", color="Assault")

```
---
The bubble chart above depicts the points recorded on arrest made in United 
States across 50 states. A bubble horizontal position shows the murder tendencies
that resulted in arrest whiles a vertical bubble position 
notes the rape rate that led to arrest in the United States. 
The size of bubble explains whether states exhibit low or high rate of these 
offences (murder or rape).The colour used allowed for transparency on the points,
with the occurrence of overlapping which suggest limitation to the number data 
points that can be plotted. 



```{r}
################ box plot####################
###using USArrest dataset#####
boxplot(Rape~UrbanPop,data=USArrests_data,xlab= "UrbanPop",ylab= "Rape",
        col= "yellow",border= "brown",
        main= "Boxplot on Murder/Rape in USA")


##plot2
```


```{r}
ggplot(USArrests_data, aes(x=Rape,y=Murder, fill=Region)) + 
  geom_boxplot()

```
---
The graps above shows different box plot, shapes and positions that shows a little bit of spread with the data been used. 
```{r}

############BAR PLOT########################
#########using USArrest dataset##########
### basic bar plot
##plot1
ggplot(USArrests_data, aes(x=Region, y=Murder)) +
  geom_bar(stat="identity")
##plot2
ggplot(USArrests_data, aes(x=Region, y=Assault)) +
  geom_bar(stat="identity")
##plot3
ggplot(USArrests_data,aes(x=UrbanPop,
                     y=Murder,col=Rape,fill=Rape))+geom_col()
##plot4
ggplot(USArrests,aes(x=UrbanPop,
                     y=Murder,
                     col=Rape,
                     fill=Assault))+geom_col()+labs(x="Population", y="Murder per 10,000",
                    title= "Murder with Urban Pop as a function: With Rape & Assault")

```

