---
title: "***MAP SPATIAL DATASETS IN R***"
author: "**CHARLES WIREDU**"
date: "**3/16/2022**"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
LOADING REQUIRED PACKAGES

```{r}
#install.packages("ggplot2")

#install.packages("sf")

#install.packages("raster")

#install.packages("rgeos")

#install.packages("cowplot")

#install.packages("googleway")

#install.packages("ggrepel")

#install.packages("ggspatial")

library(ggplot2)
library(sf)
library(raster)
library(rgeos)
library(cowplot)
library(googleway)
library(ggrepel)
library(ggspatial)
library(ggsn)
```
*LOADING DATASET AND VIEW DATA STRUCTURE*

```{r}
library(readr)
HARV_PlotLocations <- read_csv("~/Downloads/HARV_PlotLocations.csv")
View(HARV_PlotLocations)

##View of Data Structure
str(HARV_PlotLocations)

```

*DATA DESCRIPTION*

The dataframe contains 21 locations(rows) and 16 variables(attributes). The dataframe also contains columns with coordinate values namely; easting and northing (UTM coordinates) which can be demonstrated below,


*COLUMN NAMES*
```{r}
names(HARV_PlotLocations)


```
*IDENTIFYING COORDINATE VALUES*

The data frame contains several fields that are deem to contain spatial information , which includes the coordinate values. Below is a view of the first six rows of the coordinate values from the data frame.
```{r}
head(HARV_PlotLocations$easting)


head(HARV_PlotLocations$northing)
```

Also further exploration of the data frame suggest that there are column in the dataframe that provide information regarding Coordinate Reference System(CRS). The columns can be illustrated as follows;
```{r}
head(HARV_PlotLocations$geodeticDa)

head(HARV_PlotLocations$utmZone)
```

*RASTER FILE DETAILS*

With the use of the raster package in R , there is a possible conversion of the raster file to a data frame for visualization
```{r}
HARV_1 <- raster("/Users/carlkwamesarfo/Downloads/HARV_chmCrop.tif")

HARV_1

Summary(HARV_1)

```

*MAPPING POINTS AND RASTER IMAGE*
```{r}
#Call in point data, in this case a csv file with  coordinates
HARV_PlotLocations <- (read.csv("~/Downloads/HARV_PlotLocations.csv"))

#Make appropriate column headings
df <- data.frame(cbind(easting= HARV_PlotLocations$easting, northing= HARV_PlotLocations$northing, elevation= HARV_PlotLocations$elevation))

map <- raster("/Users/carlkwamesarfo/Downloads/HARV_chmCrop.tif")

#Reading points and ploting using ggplot
ggplot(df) +
geom_point(aes(easting, northing, color=elevation), size=3) +
coord_fixed(ratio= 1) +
scale_color_gradient(low= "blue", high="orange") +
theme_bw() +
  ggtitle("Map of Plot Location/Elevations")
```

```{r}
## Reading the raster image and converting to a data frame
CHM_HARV <- raster("/Users/carlkwamesarfo/Downloads/HARV_chmCrop.tif")

CHM_HARV_df <- as.data.frame(CHM_HARV, xy=TRUE)
names(CHM_HARV_df)[1] <- 'Easting'
names(CHM_HARV_df)[2] <- 'Northing'

```

```{r}
#Over lay the two data frames
ggplot(df)+
  geom_raster(data = CHM_HARV_df,aes(x=Easting,y=Northing,fill=HARV_chmCrop))+
  geom_point(aes(easting,northing,color=elevation),size=3)+
  scale_fill_gradient2(low= "blue", high="dark blue")+ 
  scale_color_gradient(low= "blue", high= "orange")+
  ggtitle("MAP-Northeast Worcester Elevation/HARV_CHM")

```

