---
title: "Data Visisualization on Covid-19 in USA"
author: "CHARLES WIREDU"
date: "5/05/2022"
output: 
   pdf_document:
     fig_caption: yes
     
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Introduction
With over 40 million confirmed cases of sever acute respiratory syndrome Coronavirus-2(SARS COV-2) infected disease called Coronavirus disease-2019 been reported  by the Center for Disease Control(CDC) in the United States,there has been a significant high mortality rate  recorded across age and sex.However regardless of these high numbers recorded in the United States, almost every single country in the world has recorded the existence of the infectious disease in their borders as revolving, making prevention of its spread a top priority for most governments, medical and political communities around the world.


The SARS-COVID was first detected in Wuhan, China in 2019, with the CDS subsequently confirming the case of its kind in the United states in January,2020. However since the outbreak of Covid-19, there have been several studies conducting visualization on rise of cases coupled with the high mortality rates associated with it. Whiles other studies tends to engage how to evident a scientific approach and how to develop significant research findings to flatten the curve.


With the evolving spread of the virus across states in the United States, some studies have predicted that a patients sex , age and pre-underlying health conditions might render patients vulnerable to either increase mortality or infection.The increasing mortality rate discovered in the era of the spread of COVID-19 on the entire population of the United States have been extensively researched in United Sates leading to key reports from CDC,Financial Times and Our World in Data.


In United States, Covid-19 posed a severe threat to the public health system, resulting in hospitals and health experts in hospitals been stretched to their maximum capacity early days of its discovery and spread.This however saw the introduction of string measures such as social distancing and face mask and consistent vaccination to curb the spread.
Moreover, there is consideration given to age , sex and prior health conditions as acting triggers for the infection and mortality associated with Covid-19.This reports aims to employ various tools to visualize the various key variables in the data set with the aim of enhancing proper communication to the target audience on covid-19 deaths with regards to whether age or sex matters.

This report is organised as follows;
               
*DATA OVERVIEW ========> DATA ANALYSIS/VISUALIZATION ========> CONCLUSION*                   


# 2. Data Overview
The data used for visualization in this report is basically provisional and relates to covid-19 death by sex and age reported to the National Center for Health Statistics in the United States.The initial data obtained for this report contained 85375 observations and 16 total variables. The R statistical software is applied to the dataset which makes provision for biographic information such as age and sex coupled state and country level information.Below is the reading and glimpse of the intial data for this report computed using the R statistical software.


*READING INTIAL DATASET*
```{r}
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(broom)
library(AICcmodavg)
library(colorspace)
library(pastecs)
library(summarytools)
library(plyr)
library(dplyr)
library(hrbrthemes)
library(viridis)
library(gganimate)
library(lubridate)
extrafont::loadfonts()
library(plotly)
Provisional_COVID_19_Deaths_by_Sex_and_Age_new_ <- read_csv("~/Desktop/Provisional_COVID-19_Deaths_by_Sex_and_Age (new).csv")
```

```{r}

```

For the purpose of this report and visualization, there are exclusion of some columns from the data set made available for this report which were are however not considered under this report. The data is however subjected to deep cleaning using the R statistical software to remove all missing values before analysis and visualization are conducted.The original data file downloaded had a size of 14.6MB in csv format. Also the initial number of observations of the data set before cleaning was implemented was 88128 coupled with 16 variables. The years recorded in the data set is 2020(start year) to 2022 (end year)
Below is an overview of how the data looked before cleaning is initiated of the data file.
```{r}

summary(Provisional_COVID_19_Deaths_by_Sex_and_Age_new_)
```

However for the purpose of the project and relevance(visualization), the initial data set is subjected to rigorous cleaning, in preparation of the data set for visualization and analysis. The initial data set downloaded contained missing values which are omitted in the process of generating a clean data set coupled with the dropping of other variables which are deem to be of no interest to the purpose this project. The cleaned data set however reports 40418 observations with 10 variables.
Below is an overview of the cleaned data set that will be used for data analysis and visualization.
```{r}
COVID_19_new <- subset(Provisional_COVID_19_Deaths_by_Sex_and_Age_new_, select = -c(`Pneumonia Deaths`,`Influenza Deaths`,Group,Year,Month,Footnote) )
covid.data<-COVID_19_new

covid19_data<-na.omit(covid.data)

covid19_data$Sex.code <- revalue(covid19_data$Sex, c("Male"="1", "Female"="2", "All Sexes"="3"))
covid19_data$Sex.code <- mapvalues(covid19_data$Sex, from = c("Male", "Female", "All Sexes"), to = c("1", "2", "3"))

summary(covid19_data)

```
The above reports the processes of cleaning the data before commencing data analysis and visualization. Also there is a summary descriptive statistics of the all the key variables that will be used for visualization.There is a report of the mean, standard deviation, median, mean, skewness and kurtosis of these variables. For instance the mean for covid-19 deaths, Pneumonia and covid-19 deaths and Total death is 658.46, 344.02 and 5079.05 respectively

## DATA SOURCE
The data for this project is publicly made available and can be accessed freely without a cost on the Center for Disease Control and Prevention website through this link; https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku. 
This data set is explored with plots for descriptive, distributions and correlations. For each plot , there is a further interpretation after writing the code.
The data set comes as a result of the detection of the first covid-19 case by the CDC in the United States and in efforts to find comprehensive data on covid-19 deaths happenings in the United States in relation to age and sex.

## DATA VARIABLES AND DEFINITONS
The final cleaned data set makes provision for 11 variables which will be implemented for data analysis and visualization in this project. the variables include the following;
```{r}
names(covid19_data)
```
The following details a brief description and definition of the variables mentioned above;

| STATE
|    The states under review in this report includes all the 50 states in the united states the occurrence of covid-19 moralities accross        the various hospitals in the country level.
| COVID-19 DEATHS
|    This variable reports the moralities recorded across the united states as a result of sever acute respiratory syndrome Corona               virus-2(SARS COV-2) infectious disease called Corona virus disease-2019 in the United States
| SEX
|    This variable also defines the gender(male or female) of individual death proportions resulting from Covid-19.
| TOTAL DEATH
|    This variable reports the total death resulting from covid-19 in the United States.
| AGE GROUP
|    This variable defines the various age groups and the related proportions affected by the covid-19 deaths in the united states.The data      set has age group ranging from under 1 to 85 years and over.
| START DATE/ END DATE
|    These two variables defines the start and end dates of the Covid-19 as reported by the CDC in the United States.
| PNUEMONIA, INFLUENZA OR COVID-19 DEATH
|    This variable defines the deaths reported  that resulted from pnuemonia, influenza or covid-19 in the United states


## DATA ANALYSIS AND VISUALIZATION
The project intend to find answers relating to Covid-19 deaths and as to whether age, sex and other factors are triggering factors contributing to the high mortality rates in the United States. As such analysis are conducted on the various key variables in the quest to find key relationships and patterns. These analysis coupled with visualizations intends to help audience understand the meaning behind the data set used for the project and also to informed their understanding on the Covid-19 death happening in the United States. Below are some analysis and visualization of the various variables used in this project;

| SEX 
|    The variable sex involves all sex, male and female.However there is proportion death relating to covid-19 death.The reports of 14629,       12308 and 13481 respectively for all sexes , female and male. 

```{r}
counts<-table(covid19_data$Sex)
```


```{r}
barplot(counts, main="Figure 1: Sex Distribution in Relation to Covid-19 in USA", col = "cyan",ylab= "Counts" )
```


The plot above shows that males have a higher representation than females, which can be attributed to the elimination of missing data values during the cleaning stage with the R statistical software

| AGE GROUPS
|   The project reports age groups <1 years, 1-4 years, 5-14years, 15-24 years, 25-34years, 35-64years, 65-74years, 75-84years and 85 and       above years in relation to covid -19 deaths occurrence in the United States. This is however explored graphically below;
```{r}

age_group <- c("All Ages","85 years and over","75-84 years","65-74 years","50-64 years","55-64 years","45-54 years","40-49 years","35-44 years","30-39 years","25-34 years","18-29 years","15-24 years","5-4 years","1-4 years", "0-17 years","Under 1 year")
count<- c("4562","3628","3778","3786","3540","3329","2475","2002","1670","1449","1317","1301","1342"," 1226","1441","1524","2048")
covid_19deaths<- c("5975708","1528042","1539635","1373203","1119107","876796","395427","253602","161737","104436","62831","33992","1326","561","13362","6241","935")

DF<- data.frame(age_group,count,covid_19deaths)
knitr::kable(DF,caption= "Age Groups with Covid-19 Deaths.")


barplot(table(covid19_data$`Age Group`),
main = "Figure 2: Age Group Distribution in Relation to Covid-19 in USA",
xlab = "Counts",
ylab = "Age Groups",
names.arg = c("All Ages","85 years and over","75-84 years","65-74 years","50-64 years","55-64 years","45-54 years","40-49 years","35-44 years","30-39 years","25-34 years","18-29 years","15-24 years","5-4 years","1-4 years", "0-17 years","Under 1 year"),
col = "darkred",
horiz = TRUE)

```

The table plot above reports the age group distribution of the data used for this reports coupled with covid 19 death in the united states. The tables further reports the covid-19 death associated with each age group affected by the pandemic. From the table it can be asserted that more deaths in the united states were reported among the aged ranging from 65 years and upwards. This however confirms high mortality rate among this age group during the peak time of the pandemic.

| COVID-19 DEATH
|   This variable details moralities associated with the infectious disease. The United States happens to be one of the countries deeply        hurt by this disease, with the country seeing high rate of moralities accross all states.This resulted in the stretch of pressure on the     health care system in the United States during its peak time. The provisional death count is based on death certificate data recieved by     the Center for Health Statistics occurring within all the 50 States. This is visually reported  coupled with other control variables        such age  and  sex using the R statistical software.
```{r}
age_data <- unique(covid19_data$`Age Group`)
age_data <- age_data[-c(3, 7, 9, 12)]
gender_data <- unique(covid19_data$Sex)
View(gender_data)

covid19_data <- covid19_data %>%
  filter(
    State!= 'Puerto Rico',
    `Age Group` %in% age_data
  )
total_deaths <- covid19_data %>%
  filter(`Total Deaths`  == 'All Ages')
#view(`Total Deaths` )
covid_new <- na.omit(covid19_data)
#names(covid_new)
#head(covid_new)


P<-ggplot(covid_new, aes(y=log(`Total Deaths`), x= log(`COVID-19 Deaths`)) ) +
  geom_point(aes(color= factor(`Age Group`)))

P+ labs(title= "Figure 3 : Covid-19 Death Against Total Deaths in US")

```

From the above visual illustration, the data shows an uphill trend towards the right portion of the graph which indicates a positive trend. That's there is a positive linear  relationship between Covid-19 deaths and total death. This can however be asserted that  an increase in covid-19 deaths makes impacts or affects the total deaths happenings in the United States with regards to the respective age groups.

| TOTAL DEATH BY AGE GROUP AND SEX
|   The CDC prior to its(covid-19) first detection in the United States had sounded a warning call that there will be more deaths recorded      across the country, resulting in State mandates across the country to curb the spread. The reported mean from the data set is 5079.05       and a standard deviation of 63841.65. Below is graphical illustration of total death from covid-19 in relation to age group and sex.

```{r}
library(rlang)
ggplot(covid_new, aes(fill=Sex, y=`Age Group`, x=`Total Deaths`)) + 
  geom_bar(position="stack", stat="identity") +
  scale_fill_viridis(discrete = T) +
  ggtitle("Figure 4: Total death across sex and age group in USA") +
  xlab("")

ggplot(covid_new,aes(x = Sex, y = `Total Deaths`,fill=`Age Group`)) + 
  geom_bar(position="dodge2", stat="identity") + 
  ggtitle("Figure 5: Total death by sex in USA") +
  facet_wrap(~Sex, nrow = 2)
```
The plots above shows the total death happenings across age group and sex(male and female).However the second also serves as visualization of total death resulting from the infectious disease(COVID-19) in a group plot in relation to sex  and age group.

| STATES AND COVID-19 DEATHS
|    All states within the United States reported fatalities occurring because of the severe acute respiratory syndrome                       Corona virus-2(SARS-COV-2) infected disease called Coronavirus-19. This was documented ranging from the hospitals to senior assisted        living homes within each and every state. This however resulted to most states such as New York imposing lock-down restrictions and state   emergency order on its residents. Below is plot to visualize states and covid-19 deaths occurrences.
```{r}
library(readxl)
 COVID19_STATES <- read_excel("~/Desktop/COVID19_STATES.xlsx")
ggplot(COVID19_STATES, aes(x=COVID19_DEATHS, y=US_State, fill=YEAR, stat = "identity")) + 
ggtitle("Figure 6: US STATES PER COVID-19 DEATHS OCCURRENCE") +
geom_boxplot()
```

| CUMMULATIVE COVID-19 CASES BY YEAR END DATES
The Covid-19 pandemic as projected by the CDC in the united states had its peak sessions with high record of cases across all the 50 states in the United States. The following plot reports the graphical illustration of the cumulative cases of covid-19 by the year end dates since it was discovered in the United States.
```{r}
datanew<- covid_new %>%
  mutate(`End Date`= mdy(`End Date`))
require(dplyr)
require(plyr)

##Animated time series 
datanew %>% group_by(`End Date`) %>%
  dplyr::summarise(count= n()) %>%
  mutate(cuml = cumsum(count)) %>%
  ggplot(aes(x=`End Date`, y= cuml)) +
  geom_line(color= "red") +
  geom_point(size= 1.5) +
  theme_bw()+
  ggtitle('Figure 7: Cummulative Cases In United States')
```
From the above plot there is a report on the cumulative number of cases with regards to the year 2021 and beyond. The plot tends to move uphill positively indicating the high record of numbers and moralities that happens afterwards.


## TABLE SUMMARY OF DATASET
Below is a table(first five) of the dataset used for the visualization in this report
```{r}
knitr::kable(head(covid_new[,3: 8]), caption= "Covid-19 Death by Age and Sex.")

```

## CONCLUSION
From the above plots, there has been exhibition of visualization of the Provisional dataset on  Covid-19 death by age and sex made available by the Center for Disease Control.These visualization ranges from bar graph to line graph to explore the dataset.In relation to figure(1), it explains the representation of sex groups used in this report coupled with mortalities associated. Also from figure(3), it can also be asserted that there is an uphill trend towards the right signifying a positive relationship between the two variables. Moreover, it can also be observed that there has been a rapid increase in the spread of the infectious disease(Covid-19) as reported by figure(7). That's there is a steep rise in the line plot towards which shows more cases recorded in the United States since its first detection. In the nutshell, the above visualization  explore all the necessary techniques with explanations that communicates to  readers of this report to enhance the understanding of covid-19 deaths in relation to sex and age.
